Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of last part of section 2 and introduction of section 3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2904 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 4e43a915a9643696e80909a94261c890ff5867ea

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters.
Moreover, by incorporating the framework of distributional modeling we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furhtermore, covariates and their possible interactions do not need to specified in advance as they are
selected automatically.

@@

In already existing generalized linear or additive approaches to circular regression modeling 
usually the location parameter $\mu$ is regressed on covariates in the general form 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\beta^{\top} z).
\end{equation}
with a circular intercept $\mu_0$, a vector of covariates $z$
with the corresponding vector of coefficients $\beta$ that are to be estimated,
and a link function $g$.
For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
on an interval of length $2\pi$ by setting 
$g(x) = 2\pi \cdot F(x)$ 
for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
constant value. Additionally to these two model variations where only one of the two distribution 
parameters depends on covariates a combination of them is introduced. In this joint model the 
two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
in alternation within an iterative optimization algorithm. 
Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
more general case by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
However, in practice the incorporated optimization procedure has shown to be very complex, already
for only one parameter being regressed on covariates and consequently even more so in the joint model.
This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
distribution with the plugged-in link function. In particular, the function is not globally concave and
possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
compression of the estimated density function.
\newline
Referring to additive models in general and hence also to their application on circular data, it has to
be considered that a proper model specification can be very challenging, particularly for a high number 
of covariates and no information on possible interactions. Moreover, the additive structure might impose 
a smooth effect even  if the true underlying effect is an abrupt shift.
\newline
One way how to deal with or even avoid the mentioned difficulties in circular regression modeling
is to employ tree-based models as presented in the following Section.

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}
The main goal of employing tree-based models to circular data is to provide a flexible and 
easy-to-apply modeling approach which allows to detect additive as well as abrupt changes while
accounting for a proper representation on the unit circle. 
%A very flexible framework should allow for detecting a wide range of 
%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
%to a circular representation should be avoided as well as any dependence on initial values.
\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
and where to split the data an angular distance measure is evaluated between the mean direction 
and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
Hence, only the location parameter is considered for splitting such that effects in other parameters 
might be missed. Therefore, we introduce a distributional approach to circular regression trees 
where a fully specified distributional model is fitted in each node of the tree which enables the 
detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
quantities.
Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
within each node is fitted without employing covariates.
%A common way of including covariates is to regress the model parameters on
%covariates. In the case of a probabilistic model for a circular response this
%corresponds to expressing the location parameter $\mu$ and the concentration
%parameter $\kappa$ of the von Mises distribution by means of the available
%covariates, e.g., in the form of a linear combination, possibly with an
%appropriate link function plugged in as in GLMs. This kind of approach is
%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
%\cite{Fisher:1993}. Apart from the methodology developed specifically for
%circular data more general approaches to regressing model parameters on
%covariates are commonly used such as for example generalized additive models
%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
%where each distribution parameter is estimated by its own generalized additive
%model (GAM). However, difficulties might arise in case of non-additive changes
%and a proper model specification can be very challenging, particularly for a
%high number of covariates and no information on possible interactions. As an
%alternative regression approach which can deal with these problems, we
%introduce a circular probabilistic tree-structured model in this section.

>>>>

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters.
By incorporating the framework of distributional modeling we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furthermore, covariates and their possible interactions do not need to specified in advance as they are
selected automatically.

@@

In most generalized linear or additive approaches to circular regression modeling 
the location parameter $\mu$ is regressed on covariates $\mathbf{z}$ in the form of 
%models employ linear combinations of the covariates to connect them to distribution 
%parameters. By plugging in a link function $g$ 
\begin{equation}
\mu = \mu_0 + g(\mathbf{\beta}^{\top} \mathbf{z})
\end{equation}
with a circular intercept $\mu_0$ and a coefficient vector $\mathbf{\beta}$; whereby the 
the additive predictor is transformed to an interval of length $2\pi$  by a link function $g$. 
As an appropriate link function typically $g(x) = 2 \cdot arctan(x)$ is employed, 
as suggested by \citet{Fisher+Lee:1992}.
While most of the proposed models estimate a constant concentration parameter, hence, assume
homogeneity, \citet{Fisher+Lee:1992} also provide a heteroscedastic version by combining two  
individual GLMs, each for one of the parameters $\mu$ and $\kappa$. 
This provides a first approach to a fully probabilistic regression model for circular 
data, however, the parameters are not regressed simultaneously on covariates as known from the 
more general framework provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}. 
However, all circular additive models incorporate the previously discussed difficulties induced by
the characteristics of the log-likelihood function and the strongly non-linear link function.
%For example, \cite{Johnson+Wehrly:1978} guarantee for the link function to return values
%on an interval of length $2\pi$ by setting 
%$g(x) = 2\pi \cdot F(x)$ 
%for a cumulative distribution function $F$ while \cite{Fisher+Lee:1992} suggest to use
%$g(x) = 2 \cdot arctan(|x|^{\lambda} \cdot sgn(x))$. 
%Most of these approaches assume homogeneity, thus, the concentration parameter $\kappa$ is estimated 
%as a constant. However, \cite{Fisher+Lee:1992} also present a model where $\kappa$ is regressed on 
%covariates employing the exponential function as a link function while $\mu$ is assumed to be a 
%constant value. Additionally to these two model variations where only one of the two distribution 
%parameters depends on covariates a combination of them is introduced. In this joint model the 
%two sets of regression coefficients employed to model $\mu$ or $\kappa$ respectively are reestimated 
%in alternation within an iterative optimization algorithm. 
%Hence, the presented theory provides a first approach to probabilistic regression modeling of circular 
%data, even though, the parameters are not regressed simultaneously on covariates as for example in a 
%more general case by generalized additive models for location, scale, and shape 
%\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
%However, in practice the incorporated optimization procedure has shown to be very complex, already
%for only one parameter being regressed on covariates and consequently even more so in the joint model.
%This is primarily due to characteristics of the log-likelihood function of the employed von Mises 
%distribution with the plugged-in link function. In particular, the function is not globally concave and
%possibly very sharp local maxima can prevent the algorithm from converging towards the true parameters.
%Furthermore, representing smooth effects on the unit circle and interpreting the corresponding coefficients
%is not straightforward, mainly owing to possibly highly nonlinear link functions. In particular, changing 
%covariates can induce rotations, however, potentially in both directions and in certain cases yielding a 
%compression of the estimated density function.
%\newline
Referring to additive models in general, it has to be considered that a proper model specification 
can be very challenging, particularly for a high number of covariates and no information on possible 
interactions. Moreover, the additive structure might impose a smooth effect even if the true underlying 
effect is an abrupt shift.

An alternative to additive models, which can deal or even 
avoid the mentioned difficulties in circular regression modeling, is to employ tree-based models. 
For the particular case of circular data, the methodology of circular regression trees and forests is
presented in the following section.

% -------------------------------------------------------------------
% Circular regression trees and forests
% -------------------------------------------------------------------
\section{Circular regression trees and forests}
\label{sec:tree_forest}

The main goal of employing tree-based models to circular data is to provide a flexible and 
easy-to-apply modeling approach which accounts for a proper representation on the unit circle.
Based on ideas of \citet{Lund:2002}, we introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models provide a very intuitive and data-driven 
alternative to commonly used circular regression models for circular data.

%A very flexible framework should allow for detecting a wide range of 
%effects, from abrupt changes to smooth transitions, while for the latter difficulties related
%to a circular representation should be avoided as well as any dependence on initial values.
%\cite{Lund:2002} introduced the concept of circular regression trees where a circular response 
%variable is regressed on linear as well as a circular covariate. For the crucial decision on how 
%and where to split the data an angular distance measure is evaluated between the mean direction 
%and each observation in a resulting subgroup in order to assess homogeneity within this subgroup. 
%Hence, only the location parameter is considered for splitting such that effects in other parameters 
%might be missed. Therefore, we introduce a distributional approach to circular regression trees 
%where a fully specified distributional model is fitted in each node of the tree which enables the 
%detecting of splits in all parameters. Moreover, not only predictions of the expected value but also 
%corresponding confidence intervals, thresholds and quantiles can be obtained among many other information
%quantities.
%Additionally, possible difficulties induced by circular rotations are avoided as the distributional model
%within each node is fitted without employing covariates.
%A common way of including covariates is to regress the model parameters on
%covariates. In the case of a probabilistic model for a circular response this
%corresponds to expressing the location parameter $\mu$ and the concentration
%parameter $\kappa$ of the von Mises distribution by means of the available
%covariates, e.g., in the form of a linear combination, possibly with an
%appropriate link function plugged in as in GLMs. This kind of approach is
%presented for example in the mixed model of \cite{Fisher+Lee:1992} and
%\cite{Fisher:1993}. Apart from the methodology developed specifically for
%circular data more general approaches to regressing model parameters on
%covariates are commonly used such as for example generalized additive models
%for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}
%where each distribution parameter is estimated by its own generalized additive
%model (GAM). However, difficulties might arise in case of non-additive changes
%and a proper model specification can be very challenging, particularly for a
%high number of covariates and no information on possible interactions. As an
%alternative regression approach which can deal with these problems, we
%introduce a circular probabilistic tree-structured model in this section.
